---
title: "basics"
output: html_document
date: "2023-10-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```


### Data
plenty of data sets are available in R; you can download them via packages.
this is nice because it's going to work on your computer as well :p


```{r data}
head(iris)
iris = iris


```

### Basic stats


```{r pressure}

mean(iris$Sepal.Width , na.rm=T) #  na.rm=T ignores missing values
median(iris$Sepal.Width)
sd(iris$Sepal.Width) 

```

You are free to use the base function but we can code this manually to better understand the formula

Here it is:


\[
\text{Sample Variance} = \frac{1}{n-1} \sum_{i=1}^{n}(x_i - \bar{x})^2
\]

Where:

- \(n\) is the number of data points.
- \(x_i\) represents each data point.
- \(\bar{x}\) is the mean of the data points.

```{r Variance manually}

variance = function(data) {
  if(!is.vector(data)) {
    stop("This function only works with a vector")
    }  #error in case data is not correctly formatted

  mean = mean(data, na.rm = T)
  
  denominator =  (length(data)-1) # (n-1) is the denominator in the variance formula.
  numerator = sum(
    (data-mean)**2
  )
  
  return(numerator/denominator) ## aplying the formula
}
```

\(\sigma\) is the square root of the variance so
to the power of 1/2 is the same as square root, 1/3 cubic root etc.

```{r Varaince}
variance(iris$Sepal.Width)
variance(iris$Sepal.Width)**(1/2) ## this is the sd.
variance(iris$Sepal.Width)**(1/2) == sd(iris$Sepal.Width) # check if we did it correctly?
```

\(\sigma\), mean & median only tell you so much about the data
graphs are always a nice tool


### plots

```{r base plots}
hist(iris$Sepal.Width)
plot(iris$Sepal.Width)
plot(iris$Sepal.Width, iris$Sepal.Length)
```

base plots are easy but don't have tons of options
ggplot has a bit of a learning curve but can provide paper worthy papers in minutes.

```{r ggplot}

ggplot( data = iris,
        map = aes(x = Sepal.Width, y = Sepal.Length)) +
  geom_point()


```
plot looks the same because we didn't use any ggplot features just yet.

```{r}

ggplot( data = iris,
        map = aes(x = Sepal.Width, y = Sepal.Length, color = Species) ) + 
  #color colors the points in their respctiv species
  geom_jitter() + theme_classic() +xlab("Width") + ylab("Length") 
#geom_jitter gets rid of the grid look whilst mostly not changing the data.
```

here we are able to see that there are 3 clusters something that wasn't clear from just the mean/median/sd

### t-tests

figures are nice but can lead to wrong interpretations - prefer formal tests
we know there is a difference per species in the iris df, can we verify this
I want to compare Setosas & Virginicas & Versicolors
I want to compare Setosas & Virginicas & Versicolors

```{r}

setosa = iris[iris$Species == "setosa", "Sepal.Width"] # here I select the df that where the species row are equal to "setosa"
virginica = iris[iris$Species == "virginica","Sepal.Width"] # same for viriginica
versiclors = iris[iris$Species == "versicolor","Sepal.Width"] # same for viriginica

```
within "[]"the space before the "," refers to the rows & after to the columns

```{r}

t.test(setosa, mu = 3)  # this tells us that the mean is significantly different from 3
t.test(virginica, versiclors) # 

```
Seems both groups are not siggnificantly different.


1. **P-value**: The p-value is a measure of the evidence against a null hypothesis. It indicates the probability of observing the data (or more extreme results) if the null hypothesis were true.

2. **T-statistic**: The t-statistic is a test statistic used in hypothesis testing. It quantifies the difference between the sample mean and the population mean, normalized by the standard error.

3. **Degrees of Freedom (df)**: Degrees of freedom represent the number of values in the final calculation of a statistic that are free to vary. In the context of t-tests, it is related to the sample size and the assumptions of the test. The formula for degrees of freedom depends on the type of t-test (e.g., independent samples or paired samples).


### Correlations

Correlation analysis is also a nice tool to asses the relationship of 2 given variables
Again base R functions exist for this

```{r}
cor(iris$Sepal.Length,iris$Sepal.Width)
cor.test(iris$Sepal.Length, iris$Sepal.Width)
```
Reminder:

\[
\text{Pearson Correlation Coefficient (r)} = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n}(x_i - \bar{x})^2 \sum_{i=1}^{n}(y_i - \bar{y})^2}}
\]

Where:

- \(n\) is the number of data points.
- \(x_i\) represents each data point in the first variable.
- \(y_i\) represents each data point in the second variable.
- \(\bar{x}\) is the mean of the first variable.
- \(\bar{y}\) is the mean of the second variable.




```{r}
calculate_correlation <- function(x, y) {
  # Calculate the mean of x and y
  mean_x <- mean(x)
  mean_y <- mean(y)

  # Calculate the sum of the products of the differences
  sum_product_diff <- sum((x - mean_x) * (y - mean_y))

  # Calculate the sum of squared differences for x and y
  sum_square_diff_x <- sum((x - mean_x)^2)
  sum_square_diff_y <- sum((y - mean_y)^2)

  # Calculate the Pearson correlation coefficient
  correlation <- sum_product_diff / sqrt(sum_square_diff_x * sum_square_diff_y)

  return(correlation)
}

calculate_correlation(iris$Sepal.Length,iris$Sepal.Width) 
cor(iris$Sepal.Length,iris$Sepal.Width)



```
### Regressions!

Of course you know about regressions. Here we are able to do mor than simply compare 2 variables as with a correlation or a t-test.



```{r}

library(stargazer)

mod1 =lm(data = iris,
   formula = Sepal.Width ~ Sepal.Length) 

mod2 =lm(data = iris,
   formula = Sepal.Width ~ Sepal.Length + Petal.Width)

mod3 =lm(data = iris,
   formula = Sepal.Width ~ Sepal.Length + Petal.Width * Petal.Length)


stargazer(mod1,mod2,mod3, type = 'text')


```
The last regression has the interaction between two terms. The interaction term should be interpreted as well as both components of it.
An interactions is exactly the same as a moderation statitically speaking!

For regressions the t-stats are used to get the p-values of the coefficeints, that is if they are significant cetibus paribus.
The F-stat tells us about the goodness of out model, the R2 how much of the Variability we are able to explain with the given model.

```{r echo = FALSE}
iv_x <- 0.3
iv_y <- 0.5
m_x <- 0.5
m_y <- 0.8
dv_x <- 0.7
dv_y <- 0.5

control_x1 <- (m_x + iv_x) / 2
control_y1 <- iv_y
control_x2 <- (m_x + dv_x) / 2
control_y2 <- dv_y


ggplot() +
  theme_void() +
  annotate("text", x = 0.5, y = 0.8, label = "Mo") +
  annotate("text", x = 0.3, y = 0.5, label = "IV") +
  annotate("text", x = 0.7, y = 0.5, label = "DV") +
  xlim(c(0.2,0.8)) + ylim(c(0.3,1)) +
  geom_segment(
    aes(x = m_x, y = m_y -.02, xend = .5, yend = .5),
    arrow = arrow(type = "closed", length = unit(0.2, "inches")),
  ) +
  geom_segment(
    aes(x = iv_x +.02, y = iv_y, xend = dv_x-.02, yend = dv_y),

    arrow = arrow(type = "closed", length = unit(0.2, "inches")),
  ) + ggtitle('Moderation')+
  labs(title = "Moderation") +
  theme(plot.title = element_text(hjust = 0.5, vjust = .8))





```

We can also do  calculate the mediation with regressions however, we now use bootstrapping to ge the Standard error (and this p-values)
I will not do this here but we did it in class using the process package. I will however show you the intuition behind the package.

Say we want to analyze the mediated effect of Petal.length on Sepal.width via Petal.Width
So:

- IV is Petal.length
- DV is Sepal.width
- Mediator is Petal.Width

```{r echo = FALSE}
iv_x <- 0.3
iv_y <- 0.5
m_x <- 0.5
m_y <- 0.8
dv_x <- 0.7
dv_y <- 0.5

control_x1 <- (m_x + iv_x) / 2
control_y1 <- iv_y
control_x2 <- (m_x + dv_x) / 2
control_y2 <- dv_y


ggplot() +
  theme_void() +
  annotate("text", x = 0.5, y = 0.8, label = "Me") +
  annotate("text", x = 0.3, y = 0.5, label = "IV") +
  annotate("text", x = 0.7, y = 0.5, label = "DV") +
  xlim(c(0.2,0.8)) + ylim(c(0.3,1)) +
  geom_curve(
    aes(x = iv_x + .02, y = iv_y, xend = m_x-.01, yend = m_y- .01),
    curvature = 0,
    arrow = arrow(type = "closed", length = unit(0.2, "inches")),
    color = "blue"
  ) +
  geom_segment(
    aes(x = m_x + .01, y = m_y -.01, xend = dv_x -.02, yend = dv_y),
    arrow = arrow(type = "closed", length = unit(0.2, "inches")),
    color = "red" # Adjust the arrow color
  ) +
  geom_segment(
    aes(x = iv_x +.02, y = iv_y, xend = dv_x-.02, yend = dv_y),

    arrow = arrow(type = "closed", length = unit(0.2, "inches")),
    color = "green" # Adjust the arrow color
  ) + ggtitle('mediation')+
   annotate("text", x = 0.4, y = 0.7, label = "a") +
  annotate("text", x = 0.6, y = 0.6, label = "b") +
  annotate("text", x = 0.5, y = 0.45, label = "c & c'") +
  labs(title = "Mediation") +
  theme(plot.title = element_text(hjust = 0.5, vjust = .8))





```

```{r}
path_c = lm(data = iris,
            Sepal.Width ~ Petal.Length)

path_a = lm(data = iris,
            Petal.Width ~ Petal.Length)

path_b = lm(data = iris,
            Sepal.Width ~ Petal.Length + Petal.Width)

indirect_effect = as.numeric(path_a$coefficients[2] * path_b$coefficients[3])
total_effect = as.numeric(path_c$coefficients[2])
direct_effect = total_effect - indirect_effect

print(paste(indirect_effect, 'is the indirect effect'))
print(paste(direct_effect, 'is the direct effect'))
print(paste(total_effect, 'is the total effect'))




```

